Computing Direct Sum Decompositions

### 2025-04-22 paper/main.tex
converted to algorithm environments

\let\oldb\b
\usepackage[dvipsnames]{xcolor}
\usepackage{notes,hyperref}
\usepackage{mathtools}
\usepackage{cleveref}
\let\b\beta

@@

  Let $P\subset \ZZ^n$ be the degrees of a set of minimal generators of $M$;
  this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
  Because $P$ is finite, we can order the elements of $P$ such that
  $\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.

\end{rem}

@@

\begin{lem}
  The eigenvalues of $\phi$ lie in $k$.
  In particular,
  for any  $\phi \in \End M$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in \End M$ has zero as an eigenvalue.
\end{lem}

Note that by ``eigenvalues of $\phi$'' we mean the roots of the characteristic polynomial $\det(\phi - t \id_M)$, which is a univariate polynomial with coefficients in $R$ and thus a priori has solutions in $t$ only in the algebraic closure of the total ring of fractions of $R$.

\begin{proof}
  Note that the expression of $\psi$ obtained in \cref{eq:Jordan} shows that the determinant of $\phi - t \id_M$ is just the product of $\mu_i -t$, and thus the eigenvalues are just the $\mu_i$, which lie in~$k$.
\end{proof}

\begin{algm}\label{alg:graded}
  Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
  \begin{enumerate}
  \item\label{item:End0} Take a random element $\phi_0$ of $[\End_R(M)]_0$, the degree zero part of $\End_R(M)$.
  \item For an eigenvalue $\mu $ of $\phi_0$, let $\phi=\phi_0-\mu \id_M$.
  \item If $\phi$ has at least two eigenvalues, let $\psi=\phi^{\mu(M)}$. By \Cref{lem:graded_algorithm}, $M\cong \im \psi \oplus \coker \psi$.  \item Repeat steps (1)--(3) for both $\im \psi$ and $\coker \psi$.
  \end{enumerate}
\end{algm}

@@

\begin{algm}\label{alg:local}
  Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
  \begin{enumerate}
    %In the graded case, take a general element $A_0$ of $[\End_R(M)]_0$, the degree zero part of $\End_R(M)$; in the local case,
  \item Take a general element $A_0$ of $\End_R(M) \setminus \m \End_R(M)$.
  \item Consider the resulting endomorphism $A$ of the $k$-vector space $M/\m M$ and find the eigenvalues of $A$.
  \item If $A$ has at least two eigenvalues, choose one eigenvalue $\lambda$, and compute a sufficiently high power $B=A^{p^{e_0}(p^{e}-1)}$ of $A$ (with the power explicitly as in the proof of \Cref{lem:jordan}). This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im B \oplus \coker B$ by \Cref{lem:idemp}.
  \item Repeat steps (1)--(3) for both $\im B$ and $\coker B$.
  \end{enumerate}
\end{algm}

@@

\begin{rem}
  Note that the above algorithm is quite sensitive to the ground field $k$, because it needs all the eigenvalues of the endomorphism $A$ of $M/\m M$ to be contained in $k$. While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field. However, the general linear combinations we take in \Cref{item:End0} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
  See \Cref{ex:elliptic} for a demonstration of the necessity of extending the base field.
\end{rem}

>>>>

\let\oldb\b
\usepackage[dvipsnames]{xcolor}
\usepackage{notes,hyperref}
\usepackage{mathtools}
\usepackage{cleveref}
\crefformat{equation}{(#2#1#3)} % omit eq.
\let\b\beta

%algorithms
\usepackage{algorithm}
\usepackage{algpseudocode}
\algdef{SE}[SUBALG]{Indent}{EndIndent}{}{\algorithmicend\ }%
\algtext*{Indent}
\algtext*{EndIndent}
\renewcommand{\algorithmicrequire}{\textbf{Input:}} % Changing "Require" to "Input"
\renewcommand{\algorithmicensure}{\textbf{Output:}} % Changing "Ensure" to "Output"

%\crefname{algocf}{alg.}{algs.}
%\Crefname{algocf}{Algorithm}{Algorithms}

@@

  Let $P\subset \ZZ^n$ be the degrees of a set of minimal generators of $M$;
  this is a poset with respect to the partial order $\a\leq \b$ if and only if $\b-\a\in \Eff R$.
  Because $P$ is finite, we can order the elements of $P$ such that
  $\deg m_1\not\leq \deg m_2\not \leq \dots\not\leq \deg m_r$.
  \mahrud{finish}
\end{rem}

@@

\begin{lem}
  The eigenvalues of $\phi$ lie in $k$. In particular, for any $\phi \in [\End M]_0$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in [\End M]_0$ has zero as an eigenvalue.
\end{lem}

\mahrud{move this above?}
By ``eigenvalues of $\phi$'' we mean the roots of the characteristic polynomial $\det(\phi - \lambda \id_M)$\mahrud{define determinant over a module, or say determinant of its cover?}, which is a univariate polynomial in $\lambda$ with coefficients in $R$ and thus a priori has solutions only in the algebraic closure of the total ring of fractions of $R$.

\begin{proof}
  The expression of $\phi$ obtained in \cref{eq:Jordan} shows that the determinant of $\phi - \lambda \id_M$ is the product of $\lambda_i - \lambda$, thus the eigenvalues are the $\lambda_i$, which all lie in $k$.
\end{proof}

\begin{algorithm}[H]
  \caption{(Indecomposable summands of a graded module over a commutative ring)}\label{alg:graded}
  \begin{algorithmic}[1]
    \smallskip
    \Require graded module $M$ over a positively graded commutative ring $R$.
    \Ensure  list $\textproc{Summands}(M)$ of indecomposable summands of $M$.
    \State Initialize a list $L$.
    \State Take a general element $\phi$ in $[\End_R(M)]_0$,
           the degree zero part of $\End_R(M)$. \label{item:End0}
    \ForAll { eigenvalues $\lambda_i$ of $\phi$ }
      \State Let $\mu_i$ be the multiplicity of $\lambda_i$.
      \State Let $\psi_i = (\phi - \lambda_i \id_M)^{\mu_i}$.
      \State Append \Call{Summands}{$\coker\psi_i$} to $L$.
    \EndFor
    \State Let $\psi$ be the composition of all $\psi_i$ above.
    \If {$\psi$ is nonzero}
      \State Append \Call{Summands}{$\im\psi$} to $L$.
    \EndIf
    \State \Return $L$.
  \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[H]
  \caption{(Indecomposable summands of a module over a commutative local ring)}\label{alg:local}
  \begin{algorithmic}[1]
    \smallskip
    % TODO: finitely generated M? noetherian R?
    \Require module $M$ over a commutative local ring $R$.
    \Ensure  list $\textproc{Summands}(M)$ of indecomposable summands of $M$.
    \State Initialize a list $L$.
    \State Take a general element $\phi$ in $\End_R(M) \setminus \m \End_R(M)$.
    \State Let $\phi_0$ be the induced endomorphism of the $k$-vector space $M/\m M$.
    % TODO: Why did we need to assume $A$ has at least two eigenvalues?
    \State Let $\lambda$ be any eigenvalue of $\phi_0$.
    \State Let $\psi = (\phi - \lambda \id_M)^{p^{e_0} (p^{e} - 1)}$
    (with the power explicitly as in the proof of \Cref{lem:jordan}).
    \State By \Cref{lem:idemp}, $\psi$ is an idempotent, hence producing a splitting of $M$.
    \If { both $\coker\psi$ and $\im\psi$ are nonzero }
      \State Append \Call{Summands}{$\coker\psi$} to $L$.
      \State Append \Call{Summands}{$\im\psi$} to $L$.
    \Else
      \State Append $M$ to $L$.
    \EndIf
    \State \Return $L$.
  \end{algorithmic}
\end{algorithm}

@@

\begin{rem}
  Note that the above algorithm is quite sensitive to the ground field $k$, because it needs all the eigenvalues of the endomorphism $A$ of $M/\m M$ to be contained in $k$. While theoretically the issue can be avoided by working over an algebraically closed ground field $\bar k$, for practical use on a computer algebra system it is better to extend $k$ to some larger finite field. However, the general linear combinations we take in Step~\ref{item:End0} of \Cref{alg:graded} should be taken with respect to the prime subfield (otherwise, as we increase the size of the finite field $k$, the eigenvalues of a general linear combination will live in higher and higher field extensions).
  See \Cref{ex:elliptic} for a demonstration of the necessity of extending the base field.
\end{rem}
